Immutability for Better State Management

### 2022-01-25 paper.tex
refined introduction and a roadMap

\section{Introduction}
In the ever growing field of computer science, and complex system design we find ourselves constantly devising new strategies for writing down a simple elegant solution, model - representation for our complex problems.
At first during the low level C, Fortran days we'd write down programs in a procedural way explicitly telling the computer what to do in each step.
then came the age of OO were we'd abstract and encapsulate data and it's behavior into Objects which served as both an api to interface with the data and a type for it and a namespace for methods that implicitly take this object as a parameter.
this methodology has dominated the industry for over 40 years which is too long for any solution in this industry.
The increasing need for distributed, stateless, fault-tolerant, concurrent, datacentric systems makes OO/imperative solutions harder to obtain and reason about. and here is where FP (functional programming ) fits in.
contrary to what you might think it's not the new kid on the block FP is as old as the first computers with lambda calculus as it's foundation FP learns a lot from it's mathematical roots in many aspects mainly with it being declarative and strict which allows for building outstanding compilers to understand it's concise declarative expressions. but in the old day computers couldn't afford the luxury of immutable data structures and so in place mutations encapsulated within a class was the most reasonable way to go about doing things.
these days memory is so cheep you can always afford to store the value and a reference to it and it's hash and a couple of almost identical copies

@@

Shadowing is a technique that could represent a changing variable. for instance, an accumulator. the technique is simple and possible in almost every programming language out there.
at it's simplest form shadowing looks like this:
\begin{lstlisting}
    // Scala  
    object Main {
        def main(args: Array[String]){
            val i = 1;
            {
                val i = 2;
                {
                    val i = 3;
                }
            }
        }  
    }    
\end{lstlisting}
But that's not really useful and even more confusing and very error prone and I'd agree shadowing in of it's self isn't really useful but it's really at the core of any recursive solution since every function come with it's own scope and 
\begin{lstlisting} 
    def factorial (n: BigDecimal):BigDecimal = {
        if (n <= 1)  1
        else n * factorial(n-1)
    }
\end{lstlisting}
Notice here n value range over \{n, ... , 1\} but it's not really changing each n deffer from the other and has it's own scope
if you run this code it'll only go so far (around  n = 9613 for this example) until you get a StackOverflowError... not good.

@@

\subsection{Laziness}
sometimes refered to as call-by-need, it's the notion that "if data won't change and functions won't neither so would results" meaning that we wouldn't perfome any operations unless they're absolutely necessary 
\lstset{style=scala}
\begin{lstlisting}
    def from(n: Int): LazyList[Int] = 
        n #:: from(n+1) 
        // here from(n+1) won't be evaluated
    def sieve (s: LazyList[Int]): LazyList[Int] = 
        s.head #:: sieve(s.tail.filter(_ % s.head != 0)) 
    // nor would sieve(s.tail.filter(_ % s.head != 0))
    val primes = sieve(from(2))

@@

\end{lstlisting}
this behavior is implemented in some languages is what's known as Function0 that is "a function without parameters" but this would be somewhat expensive since you have to pack (copy) the closure of the function with it to achieve predictable behavior contrary to just keeping references. but if the data is immutable it would make sense to only pack references this is what's refereed to as laziness

@@

%TODO: talk about referential transparency
%TODO: use these links:
%TODO: http://typeocaml.com/2015/01/02/immutable/
%TODO: https://algs4.cs.princeton.edu/23quicksort/
%TODO: https://coderscat.com/quicksort-history-and-impls/
%TODO: https://www.codingninjas.com/blog/2020/09/26/mutating-non-mutating-algorithms-in-c/
%TODO: https://elm-lang.org/news/blazing-fast-html
%TODO: https://discourse.elm-lang.org/t/can-the-compiler-skip-virtual-dom/6300
%TODO: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3136027/
%TODO:

>>>>

\section{Introduction}
In the ever growing field of computer science, and complex system design we find ourselves constantly devising new strategies for writing down a simple elegant solution, model, representation for our complex problems.
At first during the low level C, Fortran days - back when memory was expensive and scarce - we'd write down programs in a procedural way explicitly telling the computer what to do in each step and modifying data in place since we didn't afford to keep snapshots (copies, clones ...). \newline
Then when state got too big following the Object Oriented paradigm we'd abstract and encapsulate both data and meta-data along with their behavior into Objects which served as both an api to interface with the data and maybe modify it in place and a type for it to reason about it and a namespace for methods that implicitly take this object as a parameter.
this methodology has dominated the industry for over 40 years which is too long for any solution in this industry.
The increasing need for distributed, stateless, fault-tolerant, concurrent, datacentric systems makes solutions based on these old methodologies harder to obtain and reason about. and here is where Immutability and Functional Programming fits in.
contrary to what you might think it's not the new kid on the block FP is as old as the first computers with lambda calculus as it's foundation FP learns a lot from it's mathematical roots in many aspects mainly with it being declarative which allows for building outstanding compilers to understand it's concise declarative expressions. but in the old day computers couldn't afford the luxury of immutable data structures. Thus in place mutations encapsulated within an Object was the most reasonable way to go about doing things.\newline
These days memory is so cheep you can always afford to store snapshots and create new structures along the way.

@@

Shadowing is a technique that could represent a changing variable. for instance, an accumulator. the technique is simple and possible in almost every programming language out there.
at it's simplest form shadowing looks like this:
\begin{lstlisting}
    // Scala  
    object Main {
        def main(args: Array[String]){
            val i = 1;
            {
                val i = 2;
                {
                    val i = 3;
                }
            }
        }  
    }    
\end{lstlisting}
But that's not really useful and even more confusing and very error prone and I'd agree shadowing in of it's self isn't really useful but it's really at the core of any recursive solution since every function come with it's own block and 
\begin{lstlisting} 
    def factorial (n: BigDecimal):BigDecimal = {
        if (n <= 1)  1
        else n * factorial(n-1)
    }
\end{lstlisting}
Notice here n value range over \{n, ... , 1\} but it's not really changing each n deffer from the other and has it's own scope
if you run this code it'll only go so far (around  n = 9613 for this example) until you get a StackOverflowError... not good.

@@

\subsection{Laziness}
sometimes referred to as call-by-need, it's the notion that "if data won't change and functions won't neither so would results" meaning that we wouldn't perfome any operations unless they're absolutely necessary 
\lstset{style=scala}
\begin{lstlisting}
    def from(n: Int): LazyList[Int] = 
        n #:: from(n+1) 
        // here from(n+1) won't be evaluated
    def sieve (s: LazyList[Int]): LazyList[Int] = 
        s.head #:: sieve(s.tail.filter(_ % s.head != 0)) 
    // nor would sieve(s.tail.filter(_ % s.head != 0))
    val primes = sieve(from(2))

@@

\end{lstlisting}
this behavior is implemented in some languages is what's known as Function0 that is "a function without parameters" but this would be somewhat expensive since you have to pack (copy) the closure of the function with it to achieve predictable behavior contrary to just keeping references. but if the data is immutable it would make sense to only pack references this is what's referred to as laziness

@@

%TODO: talk about referential transparency
%TODO: use these links:
%TODO: http://typeocaml.com/2015/01/02/immutable/
%TODO: https://algs4.cs.princeton.edu/23quicksort/
%TODO: https://coderscat.com/quicksort-history-and-impls/
%TODO: https://www.codingninjas.com/blog/2020/09/26/mutating-non-mutating-algorithms-in-c/
%TODO: https://elm-lang.org/news/blazing-fast-html
%TODO: https://discourse.elm-lang.org/t/can-the-compiler-skip-virtual-dom/6300
%TODO: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3136027/
%TODO: StateManegment
prefered data structures
dum data classes
following strong type system
funcional
contexts static compiletime
curring dynamic clojures
compile optimizations
lazyness
transperncy substitiustion model
GC or static refs

chanllenging
modularity

bref on 
determenistic concurrency
reactiveity
modernday statless deployment
	docker or distributed or actor
	strong endpoint devices
